---
title: "Chapter 12 Notes"
subtitle: "mutate() and case_when()"
author: "Katelyn Holdeman"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
---


# Quarto Notes

In the heading, "toc" (aka table of contents) where true adds a table of contents automatically to the document. "toc-depth" with a value of 2 tells Quarto to include level 1 and 2 headings in that table of contents. Finally, "number-sections" where true adds section numbers to your headings in the rendered doc. (note: toc-depth with a value of 2 is a good default, keeps it organized without making it too crowded).

Quarto also lets you place small pieces of R code directly inside normal sentences. This is called inline code.

Inline code is useful when you want your written memo to automatically include a value computed by your R code. Quarto evaluates inline code during rendering, so the sentence stays up to date with your analysis.

In Quarto, inline code uses one pair of backticks and the engine name inside the backticks. For R, the pattern is:

backtick {r} space expression backtick

Note: "🐱 This is a good place to be curious about how Quarto works. The code is written inside a sentence, but when the report renders, Quarto runs the code and shows the result. Very cool!"

🎯 Imagine you want to write a sentence that says how many rows are in the filtered mtcars dataset. Instead of hardcoding the number, you can use inline code to compute it.

```{r}
library(dplyr)
num_rows <- mtcars %>%
  filter(cyl == 6) %>%
  nrow()
```

Memo: 🗣 The filtered dataset has `{r} num_rows` rows.

In this chapter, use inline R code when you want your memo to report a computed value such as a vector length, category count, or converted measurement.

# Load Packages

```{r}
library(dslabs)  # Data science labs package
library(dplyr)   # Data manipulation package
```

Important reminder: "🐱 Messages and warnings are often clues. A curious analyst asks what R is trying to tell them before hiding the message. Check that your packages loaded correctly."

# Vectorized Conditionals

I've used logical conditions to keep some values and drop others, or to replace selected values in place. In this chapter, I will use logical conditions in a third way: to return a result for every element. That is what functions like ifelse() and case_when() do.

Logical conditions in R can do more than select values. They can also be used to return a new vector by assigning an output to each element based on one or more rules. Functions such as ifelse() and case_when() are called vectorized conditional functions because they evaluate conditions element-by-element across a vector and return a new vector of the same length.

Now you will use conditions in a new way: not to drop or replace selected cases, but to return a result for every position. Those results are often used inside mutate() to create or modify columns.

Subsetting keeps some values and drops others. Vectorized conditionals, by contrast, keep all positions and assign each one a result. In other words, subsetting answers “What do I keep?” while vectorized conditionals answer “What value should I return for each element?”

Review section

```{r}
x <- 1:6
x > 3
```
```{r}
x[x > 3]
```
Note: 🗣 x[x > 3] subsets x to keep only values greater than 3.

"🧐 Notice: You used the same idea with data frames in Chapter 11. The filter() function also applies a logical condition, but at the row level, keeping rows that satisfy the condition and dropping the rest."

More review

Subsetting is used to replace values, called subassignment. When you assign a value to a subset of an object, you are replacing the values in those positions.

```{r}
x[x > 3] <- 99
x
```
Note: 🗣 x[x > 3] subsets x to keep only the values greater than 3, and x[x > 3] <- 99 uses logical indexing to replace those matching values with 99.

Remember:
Subsetting keeps only matching values or rows.

Subassignment changes only matching values.

Vectorized conditionals return a result for every position.

Vectorized conditional functions such as ifelse() and case_when() still test conditions element-by-element, but they do not drop nonmatching cases. Instead, they evaluate a condition for each element and return a new vector with one result for every position.

When this new vector is used inside mutate(), a function shown at the end of this chapter, it becomes a new column rather than a subset of the old one.

## ifelse()

Base R describes ifelse() as returning a value with the same shape as the test, filled from yes or no.

📜 The SYNTAX: ifelse(test, yes, no)

test – logical vector (TRUE/FALSE/NA)

yes – value to insert where test is TRUE

no – value to insert where test is FALSE.

NA – if NA, the result is NA at that position.

Goal: 🎯 Label the vector of numbers greater than 3 as “Large” and others as “Small”.

```{r}
x <- c(3, 2, 3, 6, 5)
length(x)
```
Note: "🐱 Inline R is worth practicing because it does something unusual: it lets code work inside paragrgaph sentences. I want to see whether Quarto can ask R for the length of x and place the answer directly in my memo."

🗣 `x` is a vector with `{r} length(x)` values.

Memo: I tested render and it worked, showing that "x is a vector with 5 values."

```{r}
ifelse(test = x > 3, yes = "Large", no = "Small")
```
🗣 The output positions for 3 and under are “Small” and numbers higher than 3 are “Large”.

```{r}
length(ifelse(x > 3, "Large", "Small"))
```
Note: "🗣 The result has the same length as x. Nothing was dropped. Each position got a label."

Note about curiosity: "🐱 This is the curiosity move here: I am checking whether ifelse() gave one answer back for each value I started with. I might wonder about the pattern of "Large" and "Small" later, but first I need to understand what this function returns."

💡 Tip: Read this ifelse() statement like ‘If the vector is greater than 3, then return “Large”, otherwise return “Small”.’

📖 x[x > 3] logical subsetting keeps only matches

📖 x[x > 3] <- 99 subassignment replaces only matches

⚡ ifelse(x > 3, "Large", "Small") returns a new vector with a value for every position based on the condition.

## case_when()

dplyr describes case_when() as a general vectorized if-else that evaluates rules in order and uses the first match for each element. This makes it especially useful when you have several categories to assign.

dplyr::case_when() can do exactly what ifelse() is doing.

📜 The SYNTAX: case_when(condition1 ~ value1, condition2 ~ value2, .default = default_value)

The ~ symbol means “maps to” or “yields”.

The .default argument specifies the value to return when no conditions are TRUE. If you omit .default and no conditions match, case_when() will return NA for those positions.

Goal: 🎯 Label the vector of numbers greater than 3 as “Large” and others as “Small”.

```{r}
case_when(c(3, 2, 3, 6, 5) > 3 ~ "Large", .default = "Small")
```
Note: "🗣 The output positions for 3 and under are “Small” and numbers higher than 3 are “Large”."

⚡ case_when(c(3, 2, 3, 6, 5) > 3 ~ "Large", .default = "Small") returns a new vector with a value for every position based on the condition.

Small additional note (with a lil bit of curiosity that leads into the more than two categories section next): "📖 Why do we need both ifelse() and case_when()? case_when() is usually clearer once you have 3+ categories."

## More than Two Categories



### Nested ifelse()



### Multiple Rule case_when()



# mutate()



## Column Copy



### mutate() with Arithmetic



### case_when() inside mutate()



## Overwrite Column Values


