---
title: "Chapter 12 Notes"
subtitle: "mutate() and case_when()"
author: "Katelyn Holdeman"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
---


# Quarto Notes

In the heading, "toc" (aka table of contents) where true adds a table of contents automatically to the document. "toc-depth" with a value of 2 tells Quarto to include level 1 and 2 headings in that table of contents. Finally, "number-sections" where true adds section numbers to your headings in the rendered doc. (note: toc-depth with a value of 2 is a good default, keeps it organized without making it too crowded).

Quarto also lets you place small pieces of R code directly inside normal sentences. This is called inline code.

Inline code is useful when you want your written memo to automatically include a value computed by your R code. Quarto evaluates inline code during rendering, so the sentence stays up to date with your analysis.

In Quarto, inline code uses one pair of backticks and the engine name inside the backticks. For R, the pattern is:

backtick {r} space expression backtick

Note: "🐱 This is a good place to be curious about how Quarto works. The code is written inside a sentence, but when the report renders, Quarto runs the code and shows the result. Very cool!"

🎯 Imagine you want to write a sentence that says how many rows are in the filtered mtcars dataset. Instead of hardcoding the number, you can use inline code to compute it.

```{r}
library(dplyr)
num_rows <- mtcars %>%
  filter(cyl == 6) %>%
  nrow()
```

Memo: 🗣 The filtered dataset has `{r} num_rows` rows.

In this chapter, use inline R code when you want your memo to report a computed value such as a vector length, category count, or converted measurement.

# Load Packages

```{r}
library(dslabs)  # Data science labs package
library(dplyr)   # Data manipulation package
```

Important reminder: "🐱 Messages and warnings are often clues. A curious analyst asks what R is trying to tell them before hiding the message. Check that your packages loaded correctly."

# Vectorized Conditionals



## ifelse()



## case_when()



## More than Two Categories



### Nested ifelse()



### Multiple Rule case_when()



# mutate()



## Column Copy



### mutate() with Arithmetic



### case_when() inside mutate()



## Overwrite Column Values


