---
title: "Chapter 12 Notes"
subtitle: "mutate() and case_when()"
author: "Katelyn Holdeman"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
---


# Quarto Notes

In the heading, "toc" (aka table of contents) where true adds a table of contents automatically to the document. "toc-depth" with a value of 2 tells Quarto to include level 1 and 2 headings in that table of contents. Finally, "number-sections" where true adds section numbers to your headings in the rendered doc. (note: toc-depth with a value of 2 is a good default, keeps it organized without making it too crowded).

Quarto also lets you place small pieces of R code directly inside normal sentences. This is called inline code.

Inline code is useful when you want your written memo to automatically include a value computed by your R code. Quarto evaluates inline code during rendering, so the sentence stays up to date with your analysis.

In Quarto, inline code uses one pair of backticks and the engine name inside the backticks. For R, the pattern is:

backtick {r} space expression backtick

Note: "🐱 This is a good place to be curious about how Quarto works. The code is written inside a sentence, but when the report renders, Quarto runs the code and shows the result. Very cool!"

🎯 Imagine you want to write a sentence that says how many rows are in the filtered mtcars dataset. Instead of hardcoding the number, you can use inline code to compute it.

```{r}
library(dplyr)
num_rows <- mtcars %>%
  filter(cyl == 6) %>%
  nrow()
```

Memo: 🗣 The filtered dataset has `{r} num_rows` rows.

In this chapter, use inline R code when you want your memo to report a computed value such as a vector length, category count, or converted measurement.

# Load Packages

```{r}
library(dslabs)  # Data science labs package
library(dplyr)   # Data manipulation package
```

Important reminder: "🐱 Messages and warnings are often clues. A curious analyst asks what R is trying to tell them before hiding the message. Check that your packages loaded correctly."

# Vectorized Conditionals

I've used logical conditions to keep some values and drop others, or to replace selected values in place. In this chapter, I will use logical conditions in a third way: to return a result for every element. That is what functions like ifelse() and case_when() do.

Logical conditions in R can do more than select values. They can also be used to return a new vector by assigning an output to each element based on one or more rules. Functions such as ifelse() and case_when() are called vectorized conditional functions because they evaluate conditions element-by-element across a vector and return a new vector of the same length.

Now you will use conditions in a new way: not to drop or replace selected cases, but to return a result for every position. Those results are often used inside mutate() to create or modify columns.

Subsetting keeps some values and drops others. Vectorized conditionals, by contrast, keep all positions and assign each one a result. In other words, subsetting answers “What do I keep?” while vectorized conditionals answer “What value should I return for each element?”

Review section

```{r}
x <- 1:6
x > 3
```
```{r}
x[x > 3]
```
Note: 🗣 x[x > 3] subsets x to keep only values greater than 3.

"🧐 Notice: You used the same idea with data frames in Chapter 11. The filter() function also applies a logical condition, but at the row level, keeping rows that satisfy the condition and dropping the rest."

More review

Subsetting is used to replace values, called subassignment. When you assign a value to a subset of an object, you are replacing the values in those positions.

```{r}
x[x > 3] <- 99
x
```
Note: 🗣 x[x > 3] subsets x to keep only the values greater than 3, and x[x > 3] <- 99 uses logical indexing to replace those matching values with 99.

Remember:
Subsetting keeps only matching values or rows.

Subassignment changes only matching values.

Vectorized conditionals return a result for every position.

Vectorized conditional functions such as ifelse() and case_when() still test conditions element-by-element, but they do not drop nonmatching cases. Instead, they evaluate a condition for each element and return a new vector with one result for every position.

When this new vector is used inside mutate(), a function shown at the end of this chapter, it becomes a new column rather than a subset of the old one.

## ifelse()

Base R describes ifelse() as returning a value with the same shape as the test, filled from yes or no.

📜 The SYNTAX: ifelse(test, yes, no)

test – logical vector (TRUE/FALSE/NA)

yes – value to insert where test is TRUE

no – value to insert where test is FALSE.

NA – if NA, the result is NA at that position.

Goal: 🎯 Label the vector of numbers greater than 3 as “Large” and others as “Small”.

```{r}
x <- c(3, 2, 3, 6, 5)
length(x)
```
Note: "🐱 Inline R is worth practicing because it does something unusual: it lets code work inside paragrgaph sentences. I want to see whether Quarto can ask R for the length of x and place the answer directly in my memo."

🗣 `x` is a vector with `{r} length(x)` values.

Memo: I tested render and it worked, showing that "x is a vector with 5 values."

```{r}
ifelse(test = x > 3, yes = "Large", no = "Small")
```
🗣 The output positions for 3 and under are “Small” and numbers higher than 3 are “Large”.

```{r}
length(ifelse(x > 3, "Large", "Small"))
```
Note: "🗣 The result has the same length as x. Nothing was dropped. Each position got a label."

Note about curiosity: "🐱 This is the curiosity move here: I am checking whether ifelse() gave one answer back for each value I started with. I might wonder about the pattern of "Large" and "Small" later, but first I need to understand what this function returns."

💡 Tip: Read this ifelse() statement like ‘If the vector is greater than 3, then return “Large”, otherwise return “Small”.’

📖 x[x > 3] logical subsetting keeps only matches

📖 x[x > 3] <- 99 subassignment replaces only matches

⚡ ifelse(x > 3, "Large", "Small") returns a new vector with a value for every position based on the condition.

## case_when()

dplyr describes case_when() as a general vectorized if-else that evaluates rules in order and uses the first match for each element. This makes it especially useful when you have several categories to assign.

dplyr::case_when() can do exactly what ifelse() is doing.

📜 The SYNTAX: case_when(condition1 ~ value1, condition2 ~ value2, .default = default_value)

The ~ symbol means “maps to” or “yields”.

The .default argument specifies the value to return when no conditions are TRUE. If you omit .default and no conditions match, case_when() will return NA for those positions.

Goal: 🎯 Label the vector of numbers greater than 3 as “Large” and others as “Small”.

```{r}
case_when(c(3, 2, 3, 6, 5) > 3 ~ "Large", .default = "Small")
```
Note: "🗣 The output positions for 3 and under are “Small” and numbers higher than 3 are “Large”."

⚡ case_when(c(3, 2, 3, 6, 5) > 3 ~ "Large", .default = "Small") returns a new vector with a value for every position based on the condition.

Small additional note (with a lil bit of curiosity that leads into the more than two categories section next): "📖 Why do we need both ifelse() and case_when()? case_when() is usually clearer once you have 3+ categories."

## More than Two Categories

ifelse() is designed for two categories. If you have more than two, you can nest ifelse() statements inside each other, but this quickly becomes hard to read. case_when() is designed for multiple categories and is much easier to read when you have more than two conditions.

```{r}
scores <- c(55, 65, 75, 85, 95)
```

Objective: 🎯 Re-code scores less than 60 as “Low”, scores less than 80 as “Medium”, and all others as “High”.

### Nested ifelse()

Goal: 🎯 Re-code scores less than 60 as “Low”, scores less than 80 as “Medium”, and all others as “High” using nested ifelse().

```{r}
ifelse(scores < 60, "Low",
       ifelse(scores < 80, "Medium", "High"))
```
Note: "🗣 This works, but the logic is nested inside the no argument of the first ifelse(), which becomes harder to read as the number of categories grows."

### Multiple Rule case_when()

case_when() avoids the ugly nesting that happens with repeated ifelse() calls. It is a general vectorized if-else: R checks the rules from top to bottom and uses the first rule that matches each element. That is why the order of the rules matters.

Goal: 🎯 Code scores less than 60 as "Low", scores less than 80 as "Medium", and all others as "High" using case_when().

```{r}
case_when(scores < 60 ~ "Low",
          scores < 80 ~ "Medium",
          TRUE        ~ "High")
```
Note: "🗣 This reads more like a set of classification rules: first check for "Low", then check for "Medium", and assign "High" to everything else."

Important!! ⚠️ The order of the rules matters because case_when() uses the first condition that matches.

Note about curiosity: "🐱 This is a good place to be curious and test what happens when the rules are in the wrong order. The goal is not to break the code randomly; the goal is to see how case_when() makes its decision."

```{r}
case_when(
  scores < 80 ~ "Medium",
  scores < 60 ~ "Low",
  TRUE       ~ "High"
)
```
Note: "🗣 This does not work as intended. Any score less than 60 also satisfies scores < 80, so those values are labeled "Medium" before R ever checks the "Low" rule."

💡 Tip: write rules from most specific to more general, and end with a default rule such as TRUE ~ "High".



So far, ifelse() and case_when() have returned new vectors. That is useful, but most data work happens inside data frames. Next, you will use mutate() to store those new values as columns.

# mutate()

So far, case_when() has returned a new vector of labels. In real data analysis, you usually want to store that result in a data frame as a new column. That is what mutate() does.

mutate() creates new columns that are functions of existing columns. It can also modify an existing column if you reuse the same column name.

📜 The SYNTAX:

mutate(data, new_col = expression)

Or with piping

name_data <- data %>% mutate(new_col = expression)

data is the data frame you are working with.

new_col is the name of the new column you want to create.

expression is the vector or calculation you want to store in that column.

You can think of mutate() as saying: “Take this data frame, and add a new column based on these existing columns.”

IMPORTANT!

📖 The key idea is this:

A new name on the left side of = adds a new column.

An existing name on the left side of = overwrites that column.

The right side of = tells R what values to put there.

## Column Copy

Goal: 🎯 Create a new column that is a copy of am in mtcars.

```{r}
names(mtcars)
```
```{r}
mtcars2 <- mtcars %>% 
  mutate(new_column_copy = am)
```

Note: "🗣 In a new data set called mtcars2, a new column named new_column_copy is created that is identical to the am column."

```{r}
names(mtcars2)
```

Note: "This simple example shows that mutate() can create a new column from an existing one. The right side of mutate() is for calculations or rule-based expressions on existing columns. We’ll start by using calculations inside mutate() to create a new column."

### mutate() with Arithmetic

Doing unit conversions is a common data analyst task. Here it is in action.

```{r}
measurements <- data.frame(height_in = c(60, 65, 70))
measurements
```

Note about curiosity: "🐱 This is a small but useful thing to wonder about: can Quarto pull a column name directly from R into a sentence? Let’s try it before we add a new column."

Memo: 🗣 The `measurements` data frame has one column: `{r} colnames(measurements)`.

Objective: 🎯 Create a new column called height_cm that converts inches to centimeters (1 inch = 2.54 cm).

```{r}
measurements %>%
  mutate(height_cm = height_in * 2.54)
```

Note: 🗣 mutate(height_cm = height_in * 2.54) creates a new column called height_cm. Each value in height_cm is calculated from the matching value in height_in.

### case_when() inside mutate()

The right side of mutate() can also be a rule-based expression. One common example is case_when() which lets you build a new categorical variable from an existing numeric variable.

```{r}
student_scores <- data.frame(scores = c(55, 68, 72, 83, 91))
```

Goal: 🎯 Create a new column called score_level that labels each score from scores as "Low", "Medium", or "High".

```{r}
student_scores <- student_scores %>%
  mutate(
    score_level = case_when(
      scores < 60 ~ "Low",
      scores < 80 ~ "Medium",
      TRUE        ~ "High"
    )
  )
```

🗣 mutate() adds a new column called score_level. The values in that column come from the rules inside case_when().

student_scores is the data frame.

mutate(...) adds a new column.

score_level = names that new column.

case_when(...) decides what label to return for each row based on the value in scores.

```{r}
head(student_scores)
```
🗣 `student_scores` has the columns `{r} paste(names(student_scores), collapse = " and ")`.

Note about curiosity: "🐱 Looking at this code, I might wonder how paste() works. Rather than guessing, I could check ?paste in RStudio’s Help pane. names(student_scores) gets the column names, and paste(..., collapse = " and ") combines them into a single piece of text separated by the word “and”. Notice the space before and after the and between the quotes `” and “. Those spaces are included in the output, so the column names appear with space around the word”add”."

🛠 Break Things!

🗣 `student_scores` has the columns `{r} paste(names(student_scores), collapse = "and")`.



The mutate() + case_when() duo is a powerful way to create new variables based on complex rules. You can have as many conditions as you need, and you can use existing columns in your data frame to build those rules.

In some situations, you may want to overwrite an existing column instead of creating a new one. That is also possible with mutate(), as the next section shows.

## Overwrite Column Values

mutate() will overwrite an existing column when you reuse the same column name on the left side of =. We’ll use case_when() to recode the values in the am column of mtcars from numeric to categorical labels.

Note about curiosity: "🐱 Overwriting is a place to slow down and be curious before running the code. I want to know what the original values are first, because after overwriting, those original 0 and 1 values will no longer be in the new data frame."

Goal: "🎯 Recode the am column in mtcars from 0/1 to “Automatic”/“Manual”."

```{r}
unique(mtcars$am)
```
Note: "🗣 am is a numeric column with two unique values: 0 and 1."

```{r}
cars_overwrite <- mtcars %>%
  mutate(
    am = case_when(
      am == 0 ~ "Automatic", 
      TRUE ~ "Manual")
    )
```

```{r}
unique(cars_overwrite$am)
```

🗣 The `am` column in `cars_overwrite` has the unique values: `{r} unique(cars_overwrite$am)`.

🗣 The am column has been changed from 0/1 to "Automatic"/"Manual"

IMPORTANT!!

📖 Overwrite Key Takeaway

In every example above, mutate() followed the same pattern:

the left side of = named the output column

the right side of = created its values

When the left side is a new name, R adds a new column. When the left side reused an existing name, R overwrites that column.
