---
title: "Chapter 12 Practice"
author: "Katelyn Holdeman"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
---
## Task 0

Instructions: 📦 Load the packages needed for this Practice Space: dplyr and dslabs. This setup chunk is also a good place to practice Quarto chunk controls like #| message: false and #| warning: false.

```{r}
#| message: false
#| warning: false

library(dslabs)
library(dplyr)
```

## Task 1

🧠 Concept Check

Which method returns a new vector with one result for every position?

a. Vectorized conditional functions such as ifelse() and case_when()
b. Logical subsetting using [ ]
c. Subassignment using [ ] <-
d. filter()

My answer: a. I believe that the vectorized conditional functions method returns a new vector with one result for every position.

## Task 2

🔀 Simple ifelse()

Use ifelse() to label each height in the dataset heights as "Tall" if height is greater than 68 and "Not Tall" otherwise. Store the result as height_label.

```{r}
?dslabs::heights
heights <- dslabs::heights
names(heights)
```

🗣 Add Comments: Which column contains height values in the heights data frame? Should "Tall" be assigned to the yes or no argument in the ifelse() statement? Why?

Memo: In the heights data frame, the "height" principle contains height values. (and now the word "height" doesn't look like English anymore haha). "Tall" should be assigned to the "yes" argument in the "ifelse()" statement because the yes argument applies to the values that meet the condition, or are above 68. A height above 68 inches would be considered taller than values equal to 68 inches or smaller.

```{r}
height_label <- ifelse(
  test = heights$height > 68, 
  yes = "Tall", 
  no = "Not Tall"
)
```

```{r}
table(height_label)
length(height_label)
nrow(heights)
```
Memo: I see the expected labels, "Tall" and "Not Tall." There are 508 "Not Tall" values and 542 "Tall" values.

Did ifelse() return one label for every row in heights? What evidence from your checks shows that nothing was dropped?

🪞 Reflect: I think ifelse() successfully returned a label for every row in heights. I checked the length of the height label and the label names of the "height_label" table, which showed that nothing was dropped. There were the correct label names and the same length as the original dataframe of "heights."

## Task 3

📊 Recode with mutate() and case_when()

Create mpg_recode from mtcars. Add a new column called mpg_class with three labels:

"High MPG" if mpg is greater than 25
"Low MPG" if mpg is less than 15
"Medium MPG" otherwise

```{r}
mpg_recode <- mtcars %>%
  mutate(mpg_class = case_when(
    mpg > 25 ~ "High MPG",
    mpg < 15 ~ "Low MPG",
    TRUE     ~ "Medium MPG"
  ))
```

```{r}
table(mpg_recode$mpg_class)  # expect 3 labels
head(mpg_recode)
```
Memo: I see the expected labels, "High MPG," "Low MPG," and "Medium MPG." There are 6 "High MPG" values, 5 "Low MPG" values, and 21 "Medium MPG" values.

Memo: Also, I am curious how many conditions you could do for a task like this. You would probably be using a large dataset, and it might get complex with so many conditions to code for.


## Task 4

↔︎️ Recode: Name Focus

With the dataset 🌟 stars from the dslabs package, classify each star by temperature range in a new column called temp_group. Store the result as stars_recode.

Use these rules:

temperatures less than 4000 are "Cool"
temperatures from 4000 through 7500 are "Warm"
temperatures greater than 7500 are "Hot"

```{r}
?dslabs::stars
stars <- dslabs::stars
```

```{r}
# View(stars)
colnames(stars)
summary(stars$temp)
```
Memo: The "temp" column contains temperature values in the stars data frame.

```{r}
stars_recode <- stars %>%
  mutate(
    temp_group = case_when(
      temp < 4000 ~ "Cool",
      temp <= 7500 ~ "Warm",
      TRUE          ~ "Hot"
    )
  )
```

```{r}
table(stars_recode$temp_group)  # expect 3 labels
head(stars_recode)
```

Memo: I see the expected labels, "Hot," "Cool," and "Warm." There are 35 "Cool" values, 38 "Hot" values, and 23 "Warm" values.

🪞 Reflect: The order of the temperature rules matter here because if I first wrote "temperatures less than or equal to 7500 are designated as 'Warm' temperatures," it would be misleading for low temperature values such as 0 or 5 to be labeled "Warm" instead of "Cool." It reminds me of order of operations for math. The order in which you complete operations in math problems is important, too!


## Task 5

⚖️ Unit conversion with mutate()

Create international_friendly_heights from heights by overwriting the height column. Convert the values from inches to centimeters.

💡 Tip: inches × 2.54

```{r}
heights <- dslabs::heights
head(heights)
summary(heights$height)
```

```{r}
international_friendly_heights <- heights %>%
  mutate(height = height * 2.54)
```

```{r}
head(international_friendly_heights)
summary(international_friendly_heights$height)
```

Memo: This code was able to overwrite the existing "height" column. I know it did not create a new column because there are still only 2 columns, "sex" and "height," which are the same labels as in the original "heights" data frame.

🪞 Reflect: Do the new height values look reasonable for centimeters? What changed between the original heights data frame and international_friendly_heights?

I think the new height values look reasonable for centimeters. The "height" column in the original "heights" data frame was overwritten in the new "international_friendly_heights" data frame, which converted the height column values into centimeters instead of inches. The heights themselves did not change, just the way they are measured. It makes me think of when you have a fraction like "4/6" and it's the same as saying "2/3." It's just a different way of looking at things.

## Task 6

Self-check checklist completed!
